Making Sequential Programs Deterministic

### 2019-04-05 design.tex
minor edits
macros
randoop bugs
Copy text from manual

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

@@

    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y)| : \<Det>}{\Gamma \vdash \|x : \<OrderNonDet> \ \<Map>\angles{ \<Det> \ \beta_k, \kappa_v\ \beta_v }, y : \<Det> \ \beta|}$

    \caption{Special rules for set equality and map lookup.}
\todo{In the \rulename{map precision} rule, should the \<Det> in the
  conclusion be $\beta_v$?}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List}.

>>>>

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

@@

    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y)| : \<Det> \ \beta_v}{\Gamma \vdash \|x : \<OrderNonDet> \ \<Map>\angles{ \<Det> \ \beta_k, \kappa_v\ \beta_v }, y : \<Det> \ \beta|}$

    \caption{Special rules for set equality and map lookup.}
%\todo{In the \rulename{map precision} rule, should the \<Det> in the
%  conclusion be $\beta_v$?}
    \label{fig-precision-rules}
\end{figure}

\todo{
  In the JDK, the \<equals()> method of \<Set> is annotated as

@PolyDet("up") boolean equals(@PolyDet Set<E> this, @PolyDet Object o);

While this is sound, it is imprecise.
For example, consider the code below:

    @OrderNonDet Set<@Det Integer> a; \\
    @OrderNonDet Set<@Det Integer> b; \\
    a.equals(b);

Here, the return type of \<equals()> gets the annotation \<@NonDet>.
But it would be more precise to annotate it as \<@Det>.
To achieve this precision, the checker annotates the return type
of \<equals()> called on a receiver of type \<Set> as \<@Det> if the
following conditions are true:
\begin{itemize}
    \item the receiver does not have {@code List} or its subtype as a type parameter,
    \item the argument to equals() is also an {@code @OrderNonDet Set}, and
    \item the argument to equals() also does not have {@code List} or its subtype as a
    type parameter
\end{itemize}
}

According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List}.

### 2019-04-05 java-design.tex
Continued work on case study and experiments
Tweak
Copy text from manual

@@

@@

@@

The return type of \<System.getenv>, which reads an operating system
environment variable, is always \<NonDet>.
The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
the result is \<Det> if the argument is \codeid{"line.separator"}; although
it varies by operating system, many programs including \<diff> hide those
differences from the user.
The result is also deterministic for \codeid{"path.separator"} and
\codeid{"file.separator"}, because these lead to the same behavior given
corresponding environments.
The user can specify input properties that must be passed on the \<java>
command line and thus act like inputs to the program; these are also
treated as deterministic.
The inputs to a program are also treated as \<Det>.

>>>>

We have implemented our type system for Java, in a tool named \theDeterminismChecker.

\TheDeterminismChecker requires Java version 8.

@@

\todo{
The default annotation for most unannotated types is \<@Det>.

\begin{itemize}
\item
  Formal parameter and return types default to \<@PolyDet>.
\item
  If a formal parameter/return type is an array, its component type defaults to
  \<@PolyDet>.
\item
  As an exception, the return type of a method with no unannotated or
  \<@PolyDet> formal parameters (including the receiver) defaults to
  \<@Det>.
\end{itemize}

These choices permit all possible arguments to an unannotated method.
}

@@

\todo{Explain @PolyDet("up"), "down", "use".}

@@

The return type of \<System.getenv>, which reads an operating system
environment variable, is always \<NonDet>.
The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
the result is \<Det> if the argument is \codeid{"line.separator"}; although
it varies by operating system, many programs including \<diff> hide those
differences from the user.
The result is also deterministic for \codeid{"path.separator"} and
\codeid{"file.separator"}, because these lead to the same behavior given
corresponding environments.
The user can specify input properties that must be passed on the \<java>
command line and thus act like inputs to the program; these are also
treated as deterministic.
The inputs to a program are also treated as \<Det>.  That is, the parameter
type of the \<main> method is implicitly annotated as
\<@Det String @Det []>, a deterministic array of deterministic strings.
